A Runtime-Assisted Approach To Convergence in Replicated Data Types

### 2021-11-14 introduction.tex
Figures in motivation; edits to intro

In general, however, commutativity is not a common occurrence among data
types. Most common data type definitions come with at least a pair of
operations that do not commute. For instance an \C{insert} and a \C{remove}
operations on a set do not commute if both are for the same element.
Likewise two \C{insert} operations for the same position in a list do not
commute. Such non-commutative operations translate to conflicting updates
in a distributed setting. For instance, concurrent \C{insert $k$} and
\C{remove $k$} operations for an element $k$ in a replicated set are in
conflict, and it is not clear what their cumulative result should
be\footnote{Here, the \C{insert} $i$ is assumed to have witnessed a
  \C{remove} $r'$ that is distinct from the earlier \C{remove} $r$. Both
  $i$ and $r'$ are concurrent to $r$, hence their effects are not visible
  during $r$'s execution.}. To use such a non-commutative data type in a
replicated setting, creative re-engineering of its internal representation
and algorithms is needed to turn it into a bonafide CRDT with a sensible
distributed semantics. In the case of set data type, that could involve
changing the internal representation to maintain two sets -- a set
containing the current elements and a \emph{tombstone} set that contains
every element that has ever been deleted from the set. A \C{remove
$k$} operation removes $k$ from the element set and adds it to the
tombstone set, and an \C{insert $k$} operation adds $k$ to the element set
only if it is not present in the tombstone set (i.e., only if $k$ was not
previously deleted from the set). This results in a CRDT set called
\emph{two-phase set} where an element once removed cannot be re-added.

A more complex re-engineering of the set data type could result in a set
CRDT with a different (and arguably more useful) semantics.
Sec.~\ref{sec:motivation} describes a \emph{remove-wins} CRDT set, which
allows elements to be readded, but which prioritizes \C{remove}s over
concurrent \C{insert}s. This transformation requires vector
clocks~\cite{vectorclock} to keep track of casual ordering among
\C{insert}s and \C{remove}s. In fact most CRDTs have carefully-crafted
implementations involving non-trivial technical devices like vector
clocks. For instance, Replicated Growable Array (RGA) -- a CRDT for
collaborative editing~\cite{rga}, makes use of a variant of Lamport
timestamps~\cite{lamportts}; TreeDoc, another collaborative-editing CRDT,
makes use of \emph{dense linear orders}~\cite{treedoc}; and JSON CRDT -- a
CRDT variant of the JSON storage format, implements transactional semantics
and causal dependency sets~\cite{jsoncrdt}. Such complexity makes it very
hard to reason about basic correctness properties of CRDTs, such as
convergence (formally known as \emph{strong eventual
consistency}~\cite{crdts}).  Indeed, formal verification of strong eventual
consistency for a subset of the aforementioned CRDTs required considerable
effort on behalf of multiple authors who are experts in the
field~\cite{kleppmann2017}. The need for such advanced implementation and
reasoning techniques makes the development of new CRDTs a research exercise
to be undertaken only by experts.

% hinders the widespread adoption of replicated data
% types in software development.
In this paper we propose an alternative approach to replicated data types
that addresses the problems discussed above. Our approach differs from the
CRDT approach in two fundamental ways. Firstly, we adopt a state-centric
model of replication with \emph{mergeability} of states as the defining
characteristic in contrast to CRDTs' operation-centric model characterized
by commutativity. Unlike commutativity, mergeability does not require a
full-scale re-engineering of the data type definition. Instead,
a data type only has to define a merge semantics in form a \C{merge}
function to reconcile concurrent versions of the type in presence of their
common ancestor version. Notably, \C{merge} function is left completely
unconstrained even as convergence is guaranteed on the corresponding
\emph{mergeable replicated data type} (MRDT)~\cite{mrdt}. Such strong
guarantees are possible due to the simplicity of state-based reasoning
relative to the operation- or trace-based one. Note that state-centric
model of replication has also been adapted to CRDTs, but such state-based
CRDTs require the replicated state to be organized as a lattice to
guarantee convergence -- a restriction we eliminate in our setting.

>>>>

In general, however, commutativity is not a common occurrence among
data types. Most common data type definitions come with at least a
pair of operations that do not commute. For instance an \C{add} and a
\C{remove} operations on a set do not commute if both are for the same
element.  Likewise two \C{insert} operations for the same position in
a list do not commute. Such non-commutative operations translate to
conflicting updates in a distributed setting, whose cumulative result
cannot be determined by the sequential specification of the data type
alone. To use a non-commutative data type in a replicated setting,
creative re-engineering of its internal representation and algorithms
is needed to turn it into a bonafide CRDT with a sensible distributed
semantics.
% In the case of set data type, that could involve changing the internal
% representation to maintain two sets -- a set containing the current
% elements and a \emph{tombstone} set that contains every element that
% has ever been deleted from the set. A \C{remove $k$} operation removes
% $k$ from the element set and adds it to the tombstone set, and an
% \C{insert $k$} operation adds $k$ to the element set only if it is not
% present in the tombstone set (i.e., only if $k$ was not previously
% deleted from the set). This results in a CRDT set called
% \emph{two-phase set} where an element once removed cannot be re-added. 
% A more complex re-engineering of the set data type could result in a set
% CRDT with a different (and arguably more useful) semantics.
% Sec.~\ref{sec:motivation} describes a \emph{remove-wins} CRDT set, which
% allows elements to be readded, but which prioritizes \C{remove}s over
% concurrent \C{insert}s. This transformation requires vector
% clocks~\cite{vectorclock} to keep track of casual ordering among
% \C{insert}s and \C{remove}s.
Indeed, most CRDTs have such carefully-crafted implementations that
rely on non-trivial technical devices to keep track of causal
dependencies and avoid or resolve conflicts. For instance, various
CRDT variants of the set abstract data type rely on \emph{vector
clocks} to identify conflicting \C{add}s and
\C{remove}s~\cite{zawirski-thesis, zhang}, Replicated Growable Array
(RGA) -- a CRDT for collaborative editing~\cite{rga} and JSON CRDT --
a CRDT variant of the JSON storage format, both use \emph{Lamport
timestamps}~\cite{rga, json-crdt}; and TreeDoc, another CRDT for
collaborative editing, makes use of \emph{dense linear
orders}~\cite{treedoc}. Such advanced implementations makes it hard to
reason about basic correctness properties of CRDTs, such as
convergence. Indeed, formal verification of \emph{strong eventual
consistency} (i.e., eventual convergence) for few of the
aforementioned CRDTs required considerable effort on behalf of
multiple authors who are experts in the field~\cite{kleppmann2017}.
The extraordinary effort and the expertise required to build CRDTs is
a deterrent towards using type-safe abstractions with strong
guarantees in distributed applications. To overcome this deterrent,
there is a need for an alternative approach that lets developers reuse
their sequential abstractions in a distributed setting with little to
no additional overhead of reasoning about their correctness properties
such as convergence.

% hinders the widespread adoption of replicated data
% types in software development.
In this paper we propose such an alternative approach to convergent
replicated data types.  
% The key distinguishing characteristic of
Our approach differs from the CRDT approach in
two fundamental ways. Firstly, we adopt a state-centric model of
replication with \emph{mergeability} of states as the defining
characteristic in contrast to CRDTs' operation-centric model
characterized by commutativity. Unlike commutativity, mergeability
does not require a full-scale re-engineering of the data type
definition. Instead, a data type only has to define a merge semantics
in form a \C{merge} function to reconcile concurrent versions of the
type in presence of their common ancestor version. Notably, \C{merge}
function is left completely unconstrained even as convergence is
guaranteed on the corresponding \emph{mergeable replicated data type}
(MRDT)~\cite{mrdt}. Such strong guarantees are possible due to the
simplicity of state-based reasoning relative to the operation- or
trace-based one. Note that state-centric model of replication has also
been adapted to CRDTs, but such state-based CRDTs require the
replicated state to be organized as a lattice to guarantee convergence
-- a restriction we eliminate in our setting.

### 2021-11-14 motivation.tex
Figures in motivation; edits to intro

In this section we motivate our runtime-assisted approach to
convergent RDTs with help of a set data structure.
% demonstrates the need for run-time intervention to ensure convergence.

\subsection{Add-Wins Set RDT}

Fig.~\ref{fig:ocaml-set} shows a simplified interface of a set data
type in OCaml. The interface hides a reference to a set (\C{Set.t}),
which can be updated in-place via \C{insert} and \C{remove}
operations. As such \C{Set} is an ordinary data type primarily
designed for sequential execution. Under a concurrent execution with
an asynchronously replicated state, \C{Set} would exhibit anamolous
behaviours such as those in Fig.~\ref{fig:crdt-execs}.

Fig.~\ref{fig:crdt-execs-1} shows an anamolous execution with three
replicas -- $R_1$, $R_2$, and $R_3$, all of which start with a
singleton set containing the element $e$. A client connects to the
replica $R_3$ and executes a \C{remove $e$} operation, which is then
asynchronously propagated to other replicas. Some time after applying
$R_3$'s \C{remove} at $R_2$, another client connects to $R_2$ and
re-adds $e$ by issuing an \C{insert $e$} operation. Consequently, the
state at $R_2$ is again the singleton set $\{e\}$. Replica $R_3$
however receives $R_2$'s \C{insert} ahead of $R_1$'s \C{remove},
applies them in the same order, and ends up with an empty set. The
execution thus results in divergent replica states.

Note that if set \C{insert} and \C{remove} commute, then executing
them in different order at $R_2$ and $R_3$ would not have led to
divergence. As such, \C{Set} is not a CRDT due to the admittance of
non-commutative operations. Nonetheless \C{Set} can be transformed
into a CRDT by re-engineering its interface and operations. For
instance, the anomalous execution in Fig.~\ref{fig:crdt-execs-1} can
be pre-empted by ensuring that updates are only ever applied in the
causal order.  This can be done by extending \C{Set} with vector
clocks to keep track of the causal history of each operation. A set
\C{insert} (resp. \C{remove}) would now generate an \C{Insert} (resp.
\C{Remove}) \emph{effect} tagged with the vector clock of the origin
replica. Here, the vector clock simply records the sequence number of
last operation from each replica whose effect has been received and
applied at the current replica. When an effect is received at a
replica, it is buffered until the time all its causally-preceding
effects (as captured by the tagged vector clock) have already been
received and applied. This strategy would preempt the execution in
Fig.~\ref{fig:ctrdt-execs-1} by buffering $R_2$'s \C{insert} at $R_1$
until the causally-preceding \C{remove} of $R_1$ is received and
applied. An interface for such a \emph{causally-consistent} set RDT is
shown in Fig.~\ref{fig:cc-set}.

Unfortunately, \C{Set} data type of Fig.~\ref{fig:cc-set} still admits
divergent executions due to concurrent updates.
Fig.~\ref{fig:crdt-execs-2} describes one such execution. Here,
replicas $R_1$ and $R_2$ both start with a singleton set $\{e\}$.
Two distinct clients connect to $R_1$ and $R_2$ respectively and issue
two concurrent \C{remove $e$} operations. Later, another
client connects to $R_2$ and issues an \C{insert $e$} operation. The
effects of these operations are asynchronously applied at remote
replicas as shown in the figure, resulting in the divergent states at
$R_1$ and $R_2$.

Note that, unlike the execution in Fig.~\ref{fig:crdt-execs-1}, the
conflicting operations in Fig.~\ref{crdt-execs-2}, namely $R_1$'s
\C{remove} and $R_2$'s \C{insert}, are \emph{not} causally related,
hence their relative order is not determined by the sequential
specification of the data type. Forcing a causal relationship between
them requires synchronization between \C{insert}s and \C{remove}s,
which is expensive in an asynchronous distributed setting. It
therefore becomes inevitable to ascribe semantics to concurrent
executions to restore convergence. This is done by imposing an
\emph{arbitration order} among concurrent conflicting operations,
which are otherwise unordered. For example, in
Fig.~\ref{fig:crdt-execs-2}, we might let $R_2$'s \C{insert} override
$R_1$'s \C{remove} considering that \C{insert} is re-inserting an
element after a previous remove. Ordering concurrent removes ahead
of inserts uniformly on all replicas leads to an implementation of
\C{Set} RDT where concurrent (re-)insertions consistenty \emph{win}
over removals. Such an \emph{add-wins} set is useful, for instance, to
implement a shared shopping cart where two users can concurrently
remove an item from the cart, but if one of them re-adds it then the
item should be present in the final cart\footnote{This is in fact the
semantics of Amazon's shopping cart~\cite{dynamo}}.

Extending the \C{Set} implementation from Fig.~\ref{fig:cc-set} with
\emph{add-wins} semantics is however not trivial. The suggested
approach involves tracking element-wise causal dependencies between
the \C{remove}s and \C{insert}s by maintaining a vector clock
\emph{for each element} $e$ in the set~\cite{zawirski-thesis}. The
vector clock of $e$ records the sequence number of the last \C{insert
$e$} operation from each replica whose effect has been received and
applied at the current replica. The \C{Insert} and \C{Remove} effects
on $e$ will now be tagged with $e$'s vector clock. When an \C{Remove}
effect on $e$ is received at a replica, it is applied only if the
tagged vector clock is no less than $e$'s local vector clock, i.e.,
only if the arriving \C{Remove} has seen (i.e., causally succeeds) at
least those \C{insert $e$} operations the current replica is aware of.
Otherwise the effect is simply a no-op. This strategy would result in
convergent states despite the execution in Fig.~\ref{fig:crdt-execs-2}
as $R_2$'s \C{remove} effectively becomes a no-op at $R_1$ due to
there being at least one \C{insert} operation on $e$ that it hasn't
seen. The \C{Set} interface extended with element-wise vector clocks
is shown in Fig.~\ref{fig:rw-set}. The resultant set RDT is assumed to
be correct albeit a formal proof of convergence could not be found in
the literature.

The above exercise demonstrates the considerable ingenuity and effort
involved in deriving a convergent RDT out of such simple data type as
\C{Set}. Some of the effort can be mitigated by strengthening the
underlying system model insofar as it doesn't affect the latency and
availability of the application. For instance, a system that always
delivers messages in the causal order would automatically preempt the
execution in Fig.~\ref{fig:crdt-execs-1} without the need for
additional intervention on behalf of the developer. This is a
particularly attractive proposition considering that causal
consistency can be ``bolted on'' an existing implementation of an
eventually consistent system without weakening its
guarantees~\cite{bolton}. Unfortunately, such strengthening of the
system model would deliver no benefits to the developer if they still
have to reason about fine-grained causal dependencies to guarantee
convergence, such as in the case of Fig.~\ref{fig:crdt-execs-2}. As we
observed earlier, the execution in Fig.~\ref{fig:crdt-execs-2} seems
inevitable unless every pair of \C{insert} and \C{remove} operations
are synchronized, which, regrettably, is not a practical option. The
developer therefore seems to be stuck.

Fortunately, there is a way out of this problem and the solution
requires a switch in the perspective of replication from an
operation-centric view characterized by explicit effects to a
state-centric view characterized by explicit state merges. The
motivation for this switch is the key observation that it is possible
to induce an ordering over \emph{states} of replicas even as their
operations and effects remain unordered, as long as such states are
\emph{mergeable}. The ordering is enforced by synchronizing merges in
the background while the replica-local execution progresses
unhindered. Consequently, eventual convergence can be guaranteed
\emph{without} impacting the user-perceived latency of the operations.
Moreover, working with states instead of operations frees the latter
from having to conform to algebraic laws such as commutativity.  Put
together, this means that the clients of a replicated data type can
perform any action allowed by the sequential version of the type and
immediately see the results of their actions on the local replica
while being safe in the knowledge that replica states will eventually
converge.

To demonstrate the aforementioned intuitions we reconsider the
executions from Fig.~\ref{fig:crdt-execs}, this time in the
state-centric replication model. The cornerstone of the state-centric
model is a three-way \C{merge} function that merges concurrent
versions of the state in presence of their (lowest) common ancestor
version. In our running example, the state is a value of type
\C{Set.t}, hence \C{Set.merge} function would have the type signature:
\begin{center}
\C{Set.merge} : \C{Set.t} $\rightarrow$ \C{Set.t} $\rightarrow$ \C{Set.t}
$\rightarrow$ \C{Set.t}
\end{center}
The three arguments of \C{merge} correspond to the lowest common
ancestor (LCA) version and the two concurrent versions that
independently evolved from the LCA version. The LCA is a causal
ancestor of concurrent versions, hence causal consistency is built
into the replication model. The result of set merge, intuitively, must
contain the common elements in the two concurrent versions along with
any newly added elements in either versions. Concretely:
\begin{ocaml}
  let merge s(*lca*) s1 s2 = 
              (s1 $\cap$ s2) $\cup$ (s1 - s) $\cup$ (s2 - s)
\end{ocaml}
Equipped with the above set merge function, we can now consider the
equivalent execution of Fig.~\ref{fig:crdt-execs-1} in state-centric
model. The execution is shown in Fig.~\ref{fig:mrdt-execs-1}. The
initial version on all three replicas ($v_0$) is the singleton set
$\{e\}$. Applying operations to replicas creates new versions, e.g.,
$v_1$ on $R_3$. Changes can be propagated by merging versions, e.g.,
version $v_2$ on $R_2$ is a result of merging $v_0$ and $v_1$ in
presence of their lowest common ancestor (LCA) version
$v_0$\footnote{Versions $v_0$ and $v_1$ are not concurrent as the
  former is an ancestor of the latter. Merging $v_1$ into $v_0$ is
  nonetheless possible as $v_1$ is ahead of $v_0$ in causal order. In
  Git parlance this is a \emph{fast forward} merge.}.
Likewise $v_4$ on $R_1$ is created by merging $v_3$ and $v_0$ in
presence of their LCA $v_0$. By the end of the execution, versions
$v_4$ and $v_3$ at $R_1$ and $R_2$ (resp.) have witnessed the same set
of operations, hence are in agreement.

Convergence however is not an inherent virtue of the state-centric
replication model. Fig.~\ref{fig:mrdt-exec-2} shows the state-centric
analogue of the execution in Fig.~\ref{fig:crdt-exec-2} which
diverges. Here $R_1$ and $R_2$ start with version $v_0 \,=\, \{e\}$.
$R_2$ performs a \C{remove} and and an \C{insert} making versions
$v_1$ and $v_2$ respectively. Simultaneously, $R_1$ performs a
\C{remove} to make $v_3$. Replica $R_1$ now obtains $R_2$'s changes by
merging versions $V_1$ and $v_2$ to make new versions $v_4$ and $v_5$
respectively. LCAs for these merges are $v_0$ and $v_1$. Concurrently,
$R_2$ obtains $R_1$'s changes by merging $v_3$ with $v_2$ (LCA =
$v_0$) to make $v_6$.  Now $R_1$ and $R_2$ have same set of changes
yet their final versions differ.

Note that the anamolous execution in Fig.~\ref{fig:mrdt-execs-2} could
have been avoided had the merges between $R_1$ and $R_2$ been
linearized. Fig.~\ref{fig:mrdt-execs-3} shows an execution that only
slightly differs from the one in Fig.~\ref{fig:mrdt-execs-2}. The
difference is that the merges in Fig.~\ref{fig:mrdt-execs-3} happen
linearly: first $R_1$ is merged into $R_2$ (bringing $R_1$'s
\C{remove}), then $R_2$ into $R_1$ (bringing $R_2$'s \C{remove}),
followed again by $R_2$ into $R_1$ (bringing $R_2$'s \C{insert}). As a
result of such linearization, the final versions on $R_1$ and $R_2$
converge to the singleton set $\{e\}$. Note that there exist other
linearizations of merges; for instance, $R_1 \,\rightarrow\, R_2$
merge could be ordered between the two $R_2 \rightarrow R_1$ merges.
However, all linearizations result in the same final state $\{e\}$.
Another important point to note is that only the merges are
linearized; not the entire execution. In Fig.~\ref{fig:mrdt-execs-3},
both $R_1$ and $R_2$'s \C{remove}s remove the same element $e$, which
is not possible in a linearized execution. Leaving the execution
unconstrained is crucial to satisfy the limits imposed by the CAP
theorem on a highly-available partition-tolerant distributed system.

@@

The following section formalizes the aforementioned intuitions with
help of an abstract machine that generates only the well-formed
executions.

>>>>

In this section we motivate our runtime-assisted approach to
convergent RDTs with help of a set data structure.

\begin{figure}[ht]
\centering
\begin{ocaml}
module Set: sig
  type t (* Abstract type of the set*)    
  type elt (* Abstract type of set elements *)
  val add: elt -> unit (*Add an element *)
  val remove: elt -> unit (* Remove an element *)
end
\end{ocaml}
\caption{An OCaml interface to \C{Set} abstract data type}
\label{fig:ocaml-set}
\end{figure}

\begin{figure}[ht]
  \centering
  \begin{subfigure}[t]{0.55\columnwidth}
    \includegraphics[scale=0.35]{Figures/crdt-execs-1}
    \caption{}
    \label{fig:crdt-execs-1}
  \end{subfigure}
  \begin{subfigure}[t]{0.44\columnwidth}
    \includegraphics[scale=0.35]{Figures/crdt-execs-2}
    \caption{}
    \label{fig:crdt-execs-2}
  \end{subfigure}
\caption{Anamolous executions resulting from the Asynchronous
  replication of \C{Set}. Dashed lines denote effect propagation.}
\label{fig:crdt-execs}
\end{figure}

\noindent\paragraph{From Set ADT to Set RDT} Fig.~\ref{fig:ocaml-set}
shows a simplified interface of \C{Set} (abstract) data type in OCaml.
The interface hides a reference to a set (\C{Set.t}), which can be
updated in-place via \C{add} and \C{remove} operations. \C{Set} is an
ordinary data type meant for for sequential execution. Under a
concurrent execution with an asynchronously replicated state, \C{Set}
would exhibit anamolous behaviours such as those in
Fig.~\ref{fig:crdt-execs}.

Fig.~\ref{fig:crdt-execs-1} shows an anamolous execution with three
replicas -- $R_1$, $R_2$, and $R_3$, all of which start with a
singleton set containing the element $e$. A client connects to the
replica $R_3$ and executes a \C{remove($e$)} operation, which is then
asynchronously propagated to other replicas. Some time after applying
$R_3$'s \C{remove} at $R_2$, another client connects to $R_2$ and
re-adds $e$ by issuing an \C{add($e$)} operation. Consequently, the
state at $R_2$ is again the singleton set $\{e\}$. Replica $R_3$
however receives $R_2$'s \C{add} ahead of $R_1$'s \C{remove}, applies
them in the same order, and ends up with an empty set. The execution
thus results in divergent replica states.

Note that if \C{Set.add} and \C{Set.remove} commute, then executing
them in different order at $R_2$ and $R_3$ would not have led to
divergence. As such, \C{Set} is not a CRDT due to the admittance of
non-commutative operations. Nonetheless, there exist approaches to
transform \C{Set} into a CRDT by re-engineering its interface and
operations~\cite{crdts, zawirski-thesis, zhang}. For instance, the
anomalous execution in Fig.~\ref{fig:crdt-execs-1} can be pre-empted
by ensuring that updates are only ever applied in the causal order.
This can be done by extending \C{Set} with vector clocks to keep track
of the causal history of each operation. A set \C{add} (resp.
\C{remove}) would now generate an \C{Add} (resp.  \C{Remove})
\emph{effect} tagged with the vector clock of the origin replica.
Here, the vector clock simply records the sequence number of last
operation from each replica whose effect has been received and applied
at the current replica. When an effect is received at a replica, it is
buffered until the time all its causally-preceding effects (as
captured by the tagged vector clock) have already been received and
applied. This strategy would preempt the execution in
Fig.~\ref{fig:crdt-execs-1} by buffering $R_2$'s \C{add} at $R_1$
until the causally-preceding \C{remove} of $R_1$ is received and
applied. An interface for such a \emph{causally-consistent} set RDT is
shown in Fig.~\ref{fig:cc-set}.

\begin{figure}[ht]
\centering
\begin{ocaml}
module Set: sig
  type t      type elt
  type r_id (* Replica Id *)
  (* Vector Clock *)
  type v_clock = (r_id,int) HashTable.t 
  (* Set RDT Effects *)
  type eff = Add of elt * v_clock 
           | Remove of elt * v_clock
  type buf = eff list (* pending effects *)
  (* This replica's vector clock *)
  val local_clock: v_clock 
  val add: elt -> eff
  val remove: elt -> eff
  (* Apply an effect to the local state *)
  val apply_eff: eff -> unit
end
\end{ocaml}
\caption{An OCaml interface to \C{Set} replicated data type}
\label{fig:cc-set}
\end{figure}

\noindent\paragraph{\emph{Add-Wins} Set CRDT} Unfortunately, \C{Set}
data type of Fig.~\ref{fig:cc-set} still admits divergent executions
due to concurrent updates.  Fig.~\ref{fig:crdt-execs-2} describes one
such execution. Here, replicas $R_1$ and $R_2$ both start with a
singleton set $\{e\}$.  Two distinct clients connect to $R_1$ and
$R_2$ respectively and issue two concurrent \C{remove($e$)}
operations. Later, another client connects to $R_2$ and issues an
\C{add($e$)} operation. The effects of these operations are
asynchronously applied at remote replicas as shown in the figure,
resulting in the divergent states at $R_1$ and $R_2$.

Note that, unlike the execution in Fig.~\ref{fig:crdt-execs-1}, the
conflicting operations in Fig.~\ref{fig:crdt-execs-2}, namely $R_1$'s
\C{remove} and $R_2$'s \C{add}, are \emph{not} causally related,
hence their relative order is not determined by the sequential
specification of the data type. Forcing a causal relationship between
them requires synchronization between \C{add}s and \C{remove}s,
which is expensive in an asynchronous distributed setting. It
therefore becomes inevitable to ascribe semantics to concurrent
executions to restore convergence. This is done by imposing an
\emph{arbitration order} among concurrent conflicting operations,
which are otherwise unordered. For example, in
Fig.~\ref{fig:crdt-execs-2}, we might let $R_2$'s \C{add} override
$R_1$'s \C{remove} considering that \C{add} is re-adding an
element after a previous remove. Ordering concurrent removes ahead
of adds uiniformly on all replicas leads to an implementation of
\C{Set} RDT where concurrent (re-)adds consistenty \emph{win}
over removes. Such an \emph{add-wins} set is useful, for instance, to
implement a shared shopping cart where two users can concurrently
remove an item from the cart, but if one of them re-adds it then the
item should be present in the final cart\footnote{This is in fact the
semantics of Amazon's shopping cart~\cite{dynamo}}.

Extending the \C{Set} implementation from Fig.~\ref{fig:cc-set} with
\emph{add-wins} semantics is however not trivial. The suggested
approach involves tracking element-wise causal dependencies between
the \C{remove}s and \C{add}s by maintaining a vector clock \emph{for
each element} $e$ in the set~\cite{zawirski-thesis}. The vector clock
of $e$ records the sequence number of the last \C{add($e$)} operation
from each replica whose effect has been received and applied at the
current replica. The \C{Add} and \C{Remove} effects on $e$ will now be
tagged with $e$'s vector clock. When an \C{Remove} effect on $e$ is
received at a replica, it is applied only if the tagged vector clock
is no less than $e$'s local vector clock, i.e., only if the arriving
\C{Remove} has seen (i.e., causally succeeds) at least those
\C{add($e$)} operations the current replica is aware of.  Otherwise
the effect is simply a no-op. This strategy would result in convergent
states despite the execution in Fig.~\ref{fig:crdt-execs-2} as $R_2$'s
\C{remove} effectively becomes a no-op at $R_1$ due to there being at
least one \C{add} operation on $e$ that it hasn't seen. The
\emph{add-wins} \C{Set} interface is similar to Fig.~\ref{fig:cc-set},
except that it requires an additional data structure to track
element-wise vector clocks: 
\begin{ocaml}
  val e_clock: (elt, v_clock) HashTable.t
\end{ocaml}
The resultant set RDT is assumed to be correct albeit a formal proof
of convergence could not be found in the literature.

\noindent\paragraph{The problem with the CRDT approach} The above
exercise demonstrates the considerable ingenuity and effort involved
in deriving a convergent RDT out of such simple data type as \C{Set}.
Such sophistication makes it quite hard for non-expert developers to
build, or even customize existing replicated data types to suit the
needs of their application. Some of the effort can be mitigated by
strengthening the underlying system model insofar as it doesn't affect
the latency and availability of the application. For instance, a
system that always delivers messages in the causal order would
automatically preempt the execution in Fig.~\ref{fig:crdt-execs-1}
without the need for additional intervention on behalf of the
developer. This is a particularly attractive proposition considering
that causal consistency can be ``bolted on'' an existing
implementation of an eventually consistent system without weakening
its guarantees~\cite{bolton}. Unfortunately, such strengthening of the
system model would deliver no benefits to the developer if they still
have to reason about fine-grained causal dependencies to guarantee
convergence, such as in the case of Fig.~\ref{fig:crdt-execs-2}. As we
observed earlier, the execution in Fig.~\ref{fig:crdt-execs-2} seems
inevitable unless every pair of \C{add} and \C{remove} operations are
synchronized, which, regrettably, is not a practical option. The
developer therefore seems to be stuck.

\noindent\paragraph{Key observations guiding the solution}
Fortunately, there is a way out of this problem and the solution
requires a switch in the perspective of replication from an
operation-centric view characterized by explicit effects to a
state-centric view characterized by explicit state merges. The
motivation for this switch is the key observation that it is possible
to induce an ordering over \emph{states} of replicas even as their
operations and effects remain unordered, as long as such states are
\emph{mergeable}. The ordering is enforced by synchronizing merges in
the background while the replica-local execution progresses
unhindered. Consequently, eventual convergence can be guaranteed
\emph{without} impacting the user-perceived latency of the operations.
Moreover, working with states instead of operations frees the latter
from having to conform to algebraic laws such as commutativity.  Put
together, this means that the clients of a replicated data type can
perform any action allowed by the sequential version of the type and
immediately see the results of their actions on the local replica
while being safe in the knowledge that replica states will eventually
converge.

\begin{figure}[ht]
  \centering
  \begin{subfigure}[t]{0.55\columnwidth}
    \includegraphics[scale=0.35]{Figures/mrdt-execs-1}
    \caption{}
    \label{fig:mrdt-execs-1}
  \end{subfigure}
  \begin{subfigure}[t]{0.44\columnwidth}
    \includegraphics[scale=0.35]{Figures/mrdt-execs-2}
    \caption{}
    \label{fig:mrdt-execs-2}
  \end{subfigure}
\caption{Equivalent executions of Fig.~\ref{fig:crdt-execs} in
state-centric replication model. Dashed lines now denote state merges.}
\label{fig:mrdt-execs}
\end{figure}

\noindent\paragraph{A Mergeable Replicated Set} To demonstrate the
aforementioned intuitions we reconsider the executions from
Fig.~\ref{fig:crdt-execs}, this time in the state-centric replication
model. The cornerstone of the state-centric model is a three-way
\C{merge} function that merges concurrent versions of the state in
presence of their (lowest) common ancestor version. In our running
example, the state is a value of type \C{Set.t}, hence \C{Set.merge}
function would have the type signature:
\begin{center}
\C{Set.merge} : \C{Set.t} $\rightarrow$ \C{Set.t} $\rightarrow$ \C{Set.t}
$\rightarrow$ \C{Set.t}
\end{center}
The three arguments of \C{merge} correspond to the lowest common
ancestor (LCA) version and the two concurrent versions that
independently evolved from the LCA version. The LCA is a causal
ancestor of concurrent versions, hence causal consistency is built
into the replication model. The result of set merge, intuitively, must
contain the common elements in the two concurrent versions along with
any newly added elements in either versions. Concretely:
\begin{ocaml}
  let merge s(*lca*) s1 s2 = 
              (s1 $\cap$ s2) $\cup$ (s1 - s) $\cup$ (s2 - s)
\end{ocaml}
Equipped with the above set merge function, we can now consider the
equivalent execution of Fig.~\ref{fig:crdt-execs-1} in state-centric
model. The execution is shown in Fig.~\ref{fig:mrdt-execs-1}. The
initial version on all three replicas ($v_0$) is the singleton set
$\{e\}$. Applying operations to replicas creates new versions, e.g.,
$v_1$ on $R_3$. Changes can be propagated by merging versions, e.g.,
version $v_2$ on $R_2$ is a result of merging $v_0$ and $v_1$ in
presence of their lowest common ancestor (LCA) version
$v_0$\footnote{Versions $v_0$ and $v_1$ are not concurrent as the
  former is an ancestor of the latter. Merging $v_1$ into $v_0$ is
  nonetheless possible as $v_1$ is ahead of $v_0$ in causal order. In
  Git parlance this is a \emph{fast forward} merge.}.
Likewise $v_4$ on $R_1$ is created by merging $v_3$ and $v_0$ in
presence of their LCA $v_0$. By the end of the execution, versions
$v_4$ and $v_3$ at $R_1$ and $R_2$ (resp.) have witnessed the same set
of operations, hence are in agreement.\GK{fix this.}

\noindent\paragraph{Well-formed executions and the \quark runtime}
Convergence however is not an inherent virtue of the state-centric
replication model. Fig.~\ref{fig:mrdt-execs-2} shows the state-centric
analogue of the execution in Fig.~\ref{fig:crdt-execs-2} which
diverges. Here $R_1$ and $R_2$ start with version $v_0 \,=\, \{e\}$.
$R_2$ performs a \C{remove} and and an \C{add} making versions
$v_1$ and $v_2$ respectively. Simultaneously, $R_1$ performs a
\C{remove} to make $v_3$. Replica $R_1$ now obtains $R_2$'s changes by
merging versions $V_1$ and $v_2$ to make new versions $v_4$ and $v_5$
respectively. LCAs for these merges are $v_0$ and $v_1$. Concurrently,
$R_2$ obtains $R_1$'s changes by merging $v_3$ with $v_2$ (LCA =
$v_0$) to make $v_6$.  Now $R_1$ and $R_2$ have same set of changes
yet their final versions differ.

Note that the anamolous execution in Fig.~\ref{fig:mrdt-execs-2} could
have been avoided had the merges between $R_1$ and $R_2$ been
linearized. Fig.~\ref{fig:mrdt-execs-3} shows an execution that only
slightly differs from the one in Fig.~\ref{fig:mrdt-execs-2}. The
difference is that the merges in Fig.~\ref{fig:mrdt-execs-3} happen
linearly: first $R_1$ is merged into $R_2$ (bringing $R_1$'s
\C{remove}), then $R_2$ into $R_1$ (bringing $R_2$'s \C{remove}),
followed again by $R_2$ into $R_1$ (bringing $R_2$'s \C{add}). As a
result of such linearization, the final versions on $R_1$ and $R_2$
converge to the singleton set $\{e\}$. Note that there exist other
linearizations of merges; for instance, $R_1 \,\rightarrow\, R_2$
merge could be ordered between the two $R_2 \rightarrow R_1$ merges.
However, all linearizations result in the same final state $\{e\}$.
Another important point to note is that only the merges are
linearized; not the entire execution. In Fig.~\ref{fig:mrdt-execs-3},
both $R_1$ and $R_2$'s \C{remove}s remove the same element $e$, which
is not possible in a linearized execution. Leaving the execution
unconstrained is crucial to satisfy the limits imposed by the CAP
theorem on a highly-available partition-tolerant distributed system.

@@

The following section formalizes the aforementioned intuitions with
help of an abstract machine that generates only the well-formed
executions. \GK{fix this}
